Finite Traces of Recursive Functions in SMT

### 2015-04-28 tex/main.tex
Initial work on encoding/preliminaries.

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\sorts}{\mathsf{sort}}
\newcommand{\funcs}{\mathsf{func}}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
\newcommand{\M}{\mathcal{M}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
%\newtheorem{remark}{Remark}

@@

\author {Andrew Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3} \and more?}
\authorrunning {A. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria Nancy \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

>>>>

\newcommand{\con}[1]{\mathsf{#1}}

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\M}{\mathcal{M}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{C}}

\newcommand{\ssorts}[1]{#1^\mathrm{s}}
\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{df}}
\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}

\newcommand{\pnone}{\con{none}}
\newcommand{\ptrue}{\con{true}}
\newcommand{\pfalse}{\con{false}}
\newcommand{\pol}{\con{pol}}

%\newtheorem{remark}{Remark}

\newcommand{\Bool}{\con{Bool}}
\newcommand{\ltrue}{\top}
\newcommand{\lfalse}{\bot}
\newcommand{\lite}{\con{ite}}

\newcommand{\boolop}{\oplus}
\newcommand{\forallf}[1]{\forall_{#1}}
\newcommand{\fnull}{\emptyset}
\newcommand{\farg}[1]{a_{#1}}
\newcommand{\vecfarg}[1]{\vec{a}_{#1}}
\newcommand{\fargsort}[1]{\upsilon_{#1}}

@@

\author {Andrew Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3} \and Cesare Tinelli \inst{4}}
\authorrunning {A. Reynolds \and J. C. Blanchette \and C. Tinelli}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria Nancy \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
\and
Department of Computer Science, The University of Iowa 
}

@@

\section{Related Work}

Leon, others?

\subsection{Model Finding for Quantified Formulas in SMT}

z3, cvc4

\section{Approach}

\subsection{Preliminaries}

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of sort symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean sort $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that sort (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$,
standard boolean connectives ($\wedge$, $\vee$, $\Leftrightarrow$, etc.),
and a predicate $\lite$ of type $\Bool \times \sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$.

In this paper, we associate quantified formulas with the uninterpreted functions whose semantics they define.
We will write $\forallf{f} \vec x. \varphi$ to denote a quantified formula that is defines the semantics of uninterpreted function $f$.
For consistency, we consider only quantified formulas of this form, 
and allow $f$ to be a distinguished constant $\fnull$, 
denoting that the quantified formula does not define the semantics of an uninterpreted function.
When $f$ is not $\fnull$, we will refer to $\forallf{f} \vec x. \varphi$ as a \emph{function definition} (for $f$).
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

\begin{definition}
A formula $\varphi$ is well-defined with respect to functions $S$ if it is of the form:
\begin{equation*}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{equation*}
where $f_1, \ldots, f_n$ are distinct, 
$S = \{ f_1, \ldots, f_n \}$,
and $\psi$ contains no function definitions.
\end{definition}

Given a signature $\Sigma$, 
we parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and $\sfunndefs{\Sigma}$ are the others.
In this paper, we will consider $\Sigma$-formulas that are well-defined for $\sfundefs{\Sigma}$.

We write $\terms( t )$ to denote the set of subterms in $t$.
We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$, where $f \in \sfundefs{\Sigma}$.

\subsection{Encoding Well-Defined Assumption for Recursive Functions}

In the remainder of the section, we fix a signature $\Sigma$.

The main idea behind this translation is to abstract the set of tuples that a recursive function $f$ is invoked on as a (fresh) uninterpreted sort $\fargsort{f}$.
We extend our signature $\Sigma$ as follows.
For each $f \in\sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, we associate:
\begin{enumerate}
\item[-] an uninterpreted sort $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}

The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( t^\tau, p )$ : 
 \begin{itemize}
  \item[] $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv( t_1, p )$
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y^{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z^{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the procedure, $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$.
}
\label{fig:encoding}
\end{figure}

Figure~\ref{fig:encoding} gives a procedure $\conv$ for translating a formula $t$ containing function definitions.
The procedure takes two arguments: the term $t^\tau$ to translate, and a \emph{polarity} p (either $\ptrue$, $\pfalse$, or $\pnone$).
If function definition quantifiers $\forallf{f}$ in $t$ are such that $f$ is well-defined and $\conv( t, \ptrue )$ returns $( r, \emptyset )$,
then $r$ is equivalent to $t$.

\begin{example}
Consider the formula
\end{example}

\section{Experiments}

\section{Conclusion}
\label{sec:conclusion}
